 ---
title: "Statistics II Lab - Spring 2019 - W03"
author: "Gabriel Tarriba"
date: "March 28, 2019"
output: html_document
---

```{r include = FALSE}
source("packages.r")
library(stargazer)
```

# 1. Review questions for W05-Instrumental variables
  * Why are IV useful?
  * What properties does a valid IV have? Slide 10
  * ITT and actual treatment: Compliers, always-takers, never-takers, defiers
  * What is LATE and why isn't it simply ATE?
  * What is Two-Stage Least-Squares and why is it useful?

# 2. R script for W05
  * Simulation (lines 8-48)
  * Replication of Angrist & Krueger 1991
    - What's wrong with simply regressing wage on education?
    - dplyr verbs: mutate, group_by, summarise
    - Short instructive video on this example:
    
```{r}
browseURL("https://www.youtube.com/watch?v=pI9YGSJ2qPk")
# This series is generally a very good resource to learn econometrics!
```
    - 2SLS in R: intro to ivreg
```{r}
# Super-brief tutorial, also posted on Moodle
browseURL("https://rpubs.com/wsundstrom/t_ivreg")
```
  * Voting advice applications (VAA)
    - Idea: VAA encouragement (Instrument) -> VAA use -> Voting
    - Naive OLS regression: VAA use seems to increase voting!
    - IV regression: Actually this does not seem to be the case
    
# 3. Assignment 4 - Hints!
  * 2. Search for plausible instruments - remember the conditions:
    - Relevance: Z creates variation in D
    - Exogeneity: Z is independent of any omitted variables that we would like to control for
    - Exclusion restriction: Z only affects Y via D
    - Monotonicity: D(Z=1) >= D(Z=0) for all units
    
  *3. Bert and Ernie's effect on letter recognition
    - For (b) different plot types are possible (density, boxplots, etc.)
    - For (c) you may want to use the ctable command to produce a cross-tabulation
    - In (e) you may want to compare your IV estimate to the OLS estimates (with and without controls) 
    
  *4. Simulating flawed and weak instruments
    - This one is fairly straight-forward!
    
    