---
title: "dplyr presentation"
author: "Aje Joshua Ayodele (adapted by Marina Wyss)"
output: html_document
---
##Loading the dataset _hflights_

```{r setup, eval=FALSE}
# hflights is a datframe that R loads as part of the hflights package
# install the hflights package. 
install.packages("hflights")
```

```{r, warning=FALSE, message=FALSE}
#load the hflights, load dplyr
library(hflights)
library(dplyr)

#get acquainted with the hflights dataset
?hflights
head(hflights)

#check the names of the variables
names(hflights)
```

##dplyr main functions (verbs)
1. select() - returns a subset of the columns  
2. filter() - returns a subset of the rows  
3. arrange() - reorders the rows according to single or multiple variables  
4. mutate() - add columns from existing data  
5. summarise() or summarize() - reduces each group to a single row by calculating aggregate measures  
6. group_by() - lets you define groups within your dataset  


##select()

####Print out two columns of hflights related to delay
```{r}
head(
  select(hflights, "ArrDelay", "DepDelay")
  )
```

####Print out the columns "Origin" up to "Cancelled" of hflights
```{r}
head(
  select(hflights, Origin:Cancelled)
  )

#selecting using the column numbers (indexing)
head(
  select(hflights, 14:19)
  )
```

###Helper functions for variable selection:
* starts_with("X")
* ends_with("X")
* contains("X")

####Use select() and a helper function to print out a tibble that contains just "ArrDelay" and "DepDelay" of hflights
```{r}
head(
  select(hflights, contains("delay"))
  )
```

##mutate()

####Create a new column: GroundTime. This column is the sum of TaxiIn and TaxiOut columns
```{r}
# use the pipe operator for code clarity
hflights <- hflights %>% 
  mutate(GroundTime = TaxiIn + TaxiOut)
```

##filter()

####Print out all flights in hflights that traveled 3000 or more miles
```{r}
longDistance <- hflights %>% 
  filter(Distance > 3000)

#see result
summary(longDistance$Distance)
```

####Print out all flights that departed late but arrived ahead of schedule
```{r}
filteredFlights <- hflights %>% 
  filter(DepDelay > 0, ArrDelay < 0)

#see result
head(
  select(filteredFlights, DepDelay, ArrDelay)
  )
```

####Find all of the flights that were cancelled after being delayed
```{r}
filteredFlights <- hflights %>% 
  filter(DepDelay > 0, Cancelled == 1)

#see result
head(
  select(filteredFlights, DepDelay, Cancelled)
  )
```


##arrange()

####Create a dataset of cancelled and delayed flights

```{r}
dtc <- hflights %>% 
  filter(Cancelled == 1, 
         !is.na(DepDelay))
```

####Arrange dtc by departure delays so that the shortest departure delay is at the top of the data set
```{r}
arrangedFlights <- dtc %>% 
  arrange(DepDelay)

#view result. select a few columns to view with the DepDelay column
head(
  select(arrangedFlights, DepTime, ArrTime, UniqueCarrier, DepDelay, FlightNum)
)

```

##summarize() or summarise()

####Print out a summary of hflights with a single variable, max_diversion: the longest Distance for diverted flights. 
```{r}
diverted <- hflights %>% 
  filter(Diverted == 1) %>% 
  summarise(max_diveresion = max(Distance))

diverted
```

####Create a new df containing all flights flown by American Airlines (AA). Print out a summary of aa with the following variables:
* n_flights : the total number of flights* 
* n_canc : the total number of cancelled flights* 
* avg_delay : the average arrival delay of flights whose delay is not NA*

```{r}
aa <- hflights %>% 
  filter(UniqueCarrier == "AA") %>% 
  summarise(n_flights = n(), 
            n_canc = sum(Cancelled ==1), 
            avg_delay = mean(ArrDelay, na.rm = TRUE))

aa
```

##group_by()

####Use group_by() to group hflights by UniqueCarrier. summarize() the grouped df with two summary variables:
* p_canc, the percentage of cancelled flights  
* avg_delay, the average arrival delay of flights whose delay does not equal NA  


```{r}
carrier <- hflights %>%
  group_by(UniqueCarrier) %>%
  summarize(p_canc = (mean(Cancelled)),
            avg_delay = mean(ArrDelay, na.rm = TRUE)) %>%
  arrange(avg_delay, p_canc)

carrier
```






















