---
title: "Calculating NATE and ATE"
output: html_document
---

## The effect of test prep on receiving a scholarship

In this example, treatment (test prep) was not randomized.

```{r prep, message=FALSE, warning=FALSE}
library(tidyverse)

data <- data.frame("gender" = c("male","male","female","female"),
                   "testPrep" = c("Yes","No","Yes","No"),
                   "scholarship" = c(5, 2, 18, 6),
                   "noScholarship" = c(21, 32, 24, 11))
```

### NATE for the whole sample
```{r }
everyone <- data %>% 
  group_by(testPrep) %>% 
  summarize(scholarship = sum(scholarship),
            noScholarship = sum(noScholarship)) %>% 
  mutate(shareScholarship = scholarship/(scholarship + noScholarship)) %>% 
  select(testPrep, shareScholarship) %>% 
  spread(testPrep, shareScholarship) %>% 
  mutate(NATE = Yes - No)

everyone$NATE
```

The NATE of test prep for everyone was positive, with an 18.1% increased share receiving a scholarship.

### Conditioned on gender
```{r }
byGender <- data %>% 
  mutate(shareScholarship = scholarship/(scholarship + noScholarship)) %>% 
  select(gender, testPrep, shareScholarship) %>%
  spread(testPrep, shareScholarship) %>% 
  mutate(NATE = Yes - No) %>% 
  select(gender, NATE)

byGender
```

We can see that NATE was larger for male students (13.35%) than female students (7.6%).

### ATE

We can calculate the ATE by taking a weighted average of the NATE, weighted by the proportion of subjects who are male and female.

```{r }
# NATE by gender
nateMale <- byGender %>% 
  filter(gender == "male") %>% 
  pull(NATE)

nateFemale <- byGender %>% 
  filter(gender == "female") %>% 
  pull(NATE)

# proportion of subjects male/female
genderTotals <- data %>% 
  group_by(gender) %>% 
  summarize(total = sum(scholarship) + sum(noScholarship)) %>% 
  mutate(prop = total / sum(total))

propMale <- genderTotals %>% 
  filter(gender == "male") %>% 
  pull(prop)

propFemale <- genderTotals %>% 
  filter(gender == "female") %>% 
  pull(prop)

# calculating ATE
ATE <- (propFemale * nateFemale) + (propMale * nateMale)

ATE
```

The ATE tells us that receiving test prep increases the chance of receiving a scholarship by 10.5%.