---
title: "Simulating and plotting data"
output: html_document
---

### Disclaimer
You may adapt the code for your assignments, but please use different examples! We won't be able to give you points for copies.

## Example 1: Salary, education and experience
Let's assume that salary is an additive function of education and experience:
```{r warning=FALSE, message=FALSE}
library(dplyr)
set.seed(123) # for consistent results

# experience - 1000 samples with mean of 15 and of 4 (years)
exp <- rnorm(1000, 15, 4) 

# education - 1000 samples with mean of 12 and of 2 (years)
educ <- rnorm(1000, 12, 2)

# salary - 20,000 plus 100 times the product of experience and education (euro)
salary <- 20000 + 100 * (exp*educ) 

# join the dataset
dat <- data.frame(exp, educ, salary) 
```

If we plot these variables, we can see that there's no relationship:
```{r warning=FALSE, message=FALSE}
library(ggplot2)

salaryPlot <- ggplot(dat, aes(x = exp, y = educ)) +
  geom_point() + 
  xlab("Experience") + 
  ylab("Education") +
  geom_smooth(method = 'lm')

salaryPlot
```

Now, let's control for salaries. We create a dataset with salaries around the mean:

```{r }
mean(dat$salary)

dat <- dat %>% 
  subset(salary >= 37500 & salary < 38500)
```

If we plot the relationship between experience and education again, but only for people with very similar salaries, we see something different - a negative relationship.

```{r }
meanSalaryPlot <- ggplot(dat, 
                         aes(x = exp, y = educ)) +
  geom_point() + 
  xlab("Experience") + 
  ylab("Education") +
  geom_smooth(method = 'lm')

meanSalaryPlot
```


## Example 2: Sex, height, and weight

Let's think of sex, height, and weight as mutually-related variables:

```{r }
# sex - replicate male and female 500 times each 
sex <- rep(c("male", "female"), each = 500) 

# height - mean of 165 and sd of 7, plus 7 if the observation is male
height <- rnorm(500, 165, 7) + 
  10 * as.numeric(sex == "male") 

# weight - a function of height plus a random disturbance with mean 0 and sd 7
weight <- (height - 100) + rnorm(500, 0, 7)

# join the data
dat2 <- data.frame(sex, height, weight)
```

If we plot them all together, we see a positive relationship:

```{r }
heightWeight <- ggplot(dat2, 
                       aes(x = weight, y = height)) +
  geom_point() + 
  xlab("Weight (kg)") + 
  ylab("Height (cm)") +
  geom_smooth(method = 'lm')

heightWeight
```

Now, we can distinguish between men and women, but the relationship between weight and height persists within each group.

```{r }
heightWeightGender <- ggplot(dat2, 
                             aes(x = weight, y = height, 
                                 color = sex)) +
  geom_point() + 
  xlab("Weight (kg)") + 
  ylab("Height (cm)") +
  geom_smooth(method = 'lm')

heightWeightGender
```

## Example 3: Gender, shoe size, income
```{r }
# gender - replicate male and female 500 times each
gender <- rep(c("male", "female"), each = 500) 

# shoe size - random number with mean 38 and sd 4, plus 4 if the observation is male
shoeSize <- rnorm(1000, 38, 2) + 
  4 * as.numeric(gender == "male") 

# income - a random number with mean 25 and sd 2, plus 5 if the observation is male
income = rnorm(1000, 25, 2) + 
  5 * as.numeric(gender == "male") 

dat3 <- data.frame(gender, shoeSize, income, 
                   stringsAsFactors = FALSE)
```

If we plot the relationship between shoe size and income, there appears to be a positive relationship:

```{r }
shoePlot <- ggplot(dat3, 
                   aes(x = shoeSize, y = income)) +
  geom_point() +
  geom_smooth(method = 'lm') + 
  xlab("Shoe Size") + ylab("Income")

shoePlot
```

But if we condition on gender, the relationship disappears:

```{r }
shoePlotGender <- ggplot(dat3, 
                         aes(x = shoeSize, y = income, 
                             color = gender)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  xlab("Shoe Size") + ylab("Income") 

shoePlotGender
```