---
title: "Calculating NATE and ATE"
output: html_document
---

## The effect of meth use in scholarship attainment (Also known as the meth user student's paradox (?))

You are working at the Ministry of Education of the Republic of Brillante. Professor Walter White approaches the Ministry with his new study stating that methanphetamine use leads to scholarship attainment. The results convince Minister Jessy Pinkman, who wants to move forward with meth-use encouraging legislation. You have a hunch that there may be something else behind the results, so you dive in the data. This is what you find: 

|    Groups    |  Scholarship  |  No Scholarship  | Total  |
|--------------|---------------|------------------|--------|
|   Meth consumption   |  16           |  49              | 65     |
|  No meth consumption  |  11          |  43              | 54     |


```{r prep, message=FALSE, warning=FALSE}
library(tidyverse)

data <- data.frame("Gender"=c("Female","Female","Male","Male"),
                         "Meth"=c("Yes","No","Yes","No"),
                         "Scholarship"=c(1,2,15,9),
                         "NoScholarship"=c(19,30,30,13))
```

### NATE for the whole sample


$$Proportion\,of\,scholaships_{in\,group} = \frac{Scholarship\,_{ingroup}}{Total_{ingroup}}$$
For example, for the group that did meth:
$$Proportion\,of\,scholaships_{meth\,heads} = \frac{16}{65} = 0.246$$ 
For the group that did not do meth:
$$Proportion\,of\,scholaships_{notestprep} = \frac{11}{54} = 0.204$$
In this case, if we were to look at meth use and scholarship attainment, our naive average treatment effect (NATE) would be $$NATE = 0.246 - 0.204 = 0.042$$ or in percentage notation 4.2 percentage points. **Could it be that meth use makes us more prone to obtain scholarships? Could Professor Walter White be right?**


```{r }
# A tidy approach of what we did manually

everyone <- data %>% 
  group_by(Meth) %>% #group the data by meth and non-meth use
  summarize(Scholarship = sum(Scholarship), #sum those who received scholarships
            NoScholarship = sum(NoScholarship)) %>% #sum those who did not receive scholarships
  mutate(ShareScholarship = Scholarship/(Scholarship + NoScholarship)) %>% #create a new column with the proportion
  select(Meth, ShareScholarship) %>% #selecting only column containing whether meth was used or not 
  spread(Meth, ShareScholarship) %>% #transposing data
  mutate(NATE = Yes - No) #NATE = use - non use

everyone$NATE
```

The NATE of meth use for everyone was positive. 

You manage to access the gender disaggregated data from his study:

| Groups       | Scholarship & Meth use  | No Scholarship & Meth use | Scholarship & No Meth use | No Scholarship &  No Meth use |   
|------------- | ------------- | ------------- | ------------- |------------- |
|Female        | 1             | 19            | 2             | 30           |
|Male          | 15            | 30            | 9             | 13           |
|Total         | 16            | 49            | 11            | 43           |

### Conditioned on gender

It is your hunch that gender differences are behind Professor White's results. If we follow the same logic but we take the subject's sex into account we will get the naive average treatment effect for each group:

| Sex       | Meth use | Scholarship | No Scholarship | Total |
|------------- | ------|-------------|----------------|-------|
| Male         | Yes   |    15      | 30             | 45    | 
| Male         | No    |    9        | 13             | 22    |
| Female       | Yes   |    1       | 19             | 20    | 
| Female       | No    |    2        | 30             | 32    |

$$Proportion\,of\,scholaships^{sex}_{group} = \frac{Scholarship^{sex}_{group}}{Total^{sex}_{group}} $$

Such that for females who did meth: 

$$Proportion\,of\,scholaships^{female}_{meth\,use} = \frac{Scholarship^{female}_{meth\,use}}{Total^{female}_{meth\,use}} = \frac{1}{20} = 0.05 $$
And, females who did not do meth: 

$$Proportion\,of\,scholaships^{female}_{no\,meth\,use} = \frac{Scholarship^{female}_{no\,meth\,use}}{Total^{female}_{no\,meth\,use}} = \frac{2}{32} = 0.0625 $$

Our NATE for females would be $$NATE_{females} = 0.05 - 0.0625 = -0.0125$$ or in percentage notation **-1.25 percentage points**. You can do the same for males.

```{r }
# A tidy approach for what we did manually
byGender <- data %>% 
  mutate(shareScholarship = Scholarship/(Scholarship + NoScholarship)) %>% # getting proportions
  select(Gender, Meth, shareScholarship) %>% # selecting columns containing gender, whether meth was used, and proportions
  spread(Meth, shareScholarship) %>% #transposing data
  mutate(NATE = Yes - No) %>% #creating new column containing NATE for each gender
  select(Gender, NATE) #selecting columns gender and NATE

byGender
```

We can see that the NATE is negative for both men and women. **How can it be possible that meth use has a positive effect on scholarship attainment overall, but it is negative for both men and women?**

### ATE

We can calculate the ATE by taking a weighted average of the NATE, weighted by the proportion of subjects who are male and female. We know that we have 52 females and 67 males. Hence:

$$Proportion\,of\,gender = \frac{Subjects\,of\,gender}{Total\,subjects}$$
The proportion of male subjects would be:

$$Proportion\,of\,males = \frac{Male\,subjects}{Total\,subjects} = \frac{67}{119} = 0.563$$
The proportion of female subjects would be: 
$$Proportion\,of\,females = \frac{Female\,subjects}{Total\,subjects} = \frac{52}{119} = 0.437$$

Our Average Treatment Effect would be the treatment effect for the population (think of the EU funding example), so:

$$ATE = \pi_{females}NATEfem + \pi_{males}NATEmal$$

For this example, our ATE would be:
$$ATE = 0.437(-0.0125)\, +\, 0.563(-0.0757) = -0.0481$$
```{r }
# NATE by gender
nateMale <- byGender %>% 
  filter(Gender == "Male") %>% 
  pull(NATE)

nateFemale <- byGender %>% 
  filter(Gender == "Female") %>% 
  pull(NATE)

# proportion of subjects male/female
genderTotals <- data %>% 
  group_by(Gender) %>% 
  summarize(total = sum(Scholarship) + sum(NoScholarship)) %>% 
  mutate(prop = total / sum(total))

propMale <- genderTotals %>% 
  filter(Gender == "Male") %>% 
  pull(prop)

propFemale <- genderTotals %>% 
  filter(Gender == "Female") %>% 
  pull(prop)

# calculating ATE
ATE <- (propFemale * nateFemale) + (propMale * nateMale)

ATE
```

The ATE tells us that the effect is actually negative. It appears that the students' sex is driving both meth use and scholarship attainment ie. a confounder. Hence, when we only look at the relationship between meth use and scholarship attainment, we see a positive "effect". Once we account (control) for the confounder, we can see the true relationship.

**Note: Though this example may be more or less intuitive, (we do not expect meth use to have a positive effect in general), the policy world we navigate may have manifestations of the same phenomenon on a less intuitive manner. It is up to us, policy-makers, policy-analysts, scholars, data scientists, etc., to uncover the true mechanisms behind these relationships.**