---
title: "Simulating and plotting data"
output: html_document
---

### Disclaimer
You may adapt the code for your assignments, but please use different examples! We won't be able to give you points for copies.

```{r warning=FALSE, message=FALSE}
library(dplyr) # data wrangling
library(ggplot2) # plots
set.seed(123) # for consistent results
```

## Example 1: Salary, education and experience
Let's assume that salary is an additive function of education and experience:

```{r }
# experience - 1000 samples with mean of 15 and of 4 (years)
exp <- rnorm(1000, 15, 4) 

# education - 1000 samples with mean of 12 and of 2 (years)
educ <- rnorm(1000, 12, 2)

# salary - 20,000 plus 100 times the product of experience and education (euro)
salary <- 20000 + 100 * (exp*educ) 

# join the dataset
dat <- data.frame(exp, educ, salary) 
```

If we plot these variables, we can see that there's no relationship:
```{r }
salaryPlot <- ggplot(dat, aes(x = exp, y = educ)) +
  geom_point() + 
  xlab("Experience") + 
  ylab("Education") +
  geom_smooth(method = "lm", se = FALSE)

salaryPlot
```

Now, let's control for salaries. We create a dataset with salaries around the mean:

```{r }
mean(dat$salary)

dat <- dat %>% 
  subset(salary >= 37500 & salary < 38500)
```

If we plot the relationship between experience and education again, but only for people with very similar salaries, we see something different - a negative relationship between experience and education.

```{r }
meanSalaryPlot <- ggplot(dat, 
                         aes(x = exp, y = educ)) +
  geom_point() + 
  xlab("Experience") + 
  ylab("Education") +
  geom_smooth(method = "lm", se = FALSE)

meanSalaryPlot
```

## Example 2: Celebrity, beauty, and talent
```{r }
# beauty - 1000 observations with mean 5 and sd 1.5 (arbitrary scale)
beauty <- rnorm(1000, 5, 1.5)

# talent - 1000 observations with mean 3 and sd 1 (arbitrary scale)
talent <- rnorm(1000, 3, 1)

# celebrity - binary
celebrity <-  ifelse(beauty + talent > 8, "celeb" ,"not_celeb") 

dat <- data.frame(beauty, talent, celebrity, 
                  stringsAsFactors = FALSE)
```

Again, without conditioning, we can't see a relationship.

```{r }
celebPlot <- ggplot(dat, 
                    aes(x = beauty, y = talent)) +
    geom_point() +
  xlab("Beauty") + 
  ylab("Talent") +
  geom_smooth(method = "lm", se = FALSE)

celebPlot
```

What happens if we condition on celebrity?

```{r }
# we could either create a new data set with only celebrities:
celebData <- dat %>% 
  filter(celebrity == "celeb")

celebPlot2 <- ggplot(celebData, 
                     aes(x = beauty, y = talent)) +
    geom_point() +
  xlab("Beauty") + 
  ylab("Talent") +
  geom_smooth(method = "lm", se = FALSE)

celebPlot2

# or add celebrity as a color aesthetic:
celebPlot3 <- ggplot(dat, 
                     aes(x = beauty, y = talent,
                        col = celebrity)) +
    geom_point() +
  xlab("Beauty") + 
  ylab("Talent") +
  geom_smooth(method = "lm", se = FALSE)

celebPlot3
```

We can now see a negative relationship between talent and beauty.

## Example 3: Gender, shoe size, income
```{r }
# gender - replicate male and female 500 times each
gender <- rep(c("male", "female"), each = 500) 

# shoe size - random number with mean 38 and sd 4, plus 4 if the observation is male
shoeSize <- rnorm(1000, 38, 2) + 
  4 * as.numeric(gender == "male") 

# income - a random number with mean 25 and sd 2, plus 5 if the observation is male
income = rnorm(1000, 25, 2) + 
  5 * as.numeric(gender == "male") 

dat3 <- data.frame(gender, shoeSize, income, 
                   stringsAsFactors = FALSE)
```

If we plot the relationship between shoe size and income, there appears to be a positive relationship:

```{r }
shoePlot <- ggplot(dat3, 
                   aes(x = shoeSize, y = income)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  xlab("Shoe Size") + 
  ylab("Income")

shoePlot
```

But if we condition on gender, the relationship disappears:

```{r }
shoePlotGender <- ggplot(dat3, 
                         aes(x = shoeSize, y = income, 
                             color = gender)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Shoe Size") + 
  ylab("Income") 

shoePlotGender
```