---
title: "Matching and Logistic Regression"
output: html_document
---
## The Effect of Catholic School on Student Achievement

Example inspired by Simon Ejdemyr: https://sejdemyr.github.io/r-tutorials/statistics/tutorial8.html

#### Reference links:
1. MatchIt: https://cran.r-project.org/web/packages/MatchIt/vignettes/matchit.pdf
2. Cobalt (optional library for matching plots and extra features): https://cran.r-project.org/web/packages/cobalt/vignettes/cobalt_A0_basic_use.html
3. kableExtra (for formatting data frames): https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html
4. Stargazer (for formatting model outputs): https://www.jakeruss.com/cheatsheets/stargazer/
5. Video overview of matching concepts: https://fr.coursera.org/lecture/crash-course-in-causality/overview-of-matching-JQfPC

```{r warning=FALSE, message=FALSE}
library(dplyr) # for data wrangling
library(ggplot2) # for creating plots
library(stargazer) # for formatting model output
library(kableExtra) # for formatting data frames
library(MatchIt) # for matching

set.seed(123) # for consistent results
```

### Data Prep and EDA

First, let's read in the data. We will remove a few variables that won't be useful for our analysis. Then, we will compare the students *pre-matching* on their standardized math scores (c5r2mtsc_std).

```{r warning=FALSE, message=FALSE}
ecls <- read.csv("data/ecls.csv")

ecls <- ecls %>% 
  select(-childid, -race, -w3daded,
         -w3momed, -w3inccat)

ecls %>%
  group_by(catholic) %>%
  summarise(n_students = n(),
            mean_math = mean(c5r2mtsc_std),
            std_error = sd(c5r2mtsc_std) / sqrt(n_students)) %>% 
  round(3) %>% # round the results
  kable() %>% # create kable table
  kable_styling() # view kable table
```

We can see that we have many more students that did not attend Catholic school than those who did, and the Catholic school students have a mean math score that is higher. 

### NATE

First, let's calculate the NATE of Catholic school on student achievement (without adjusting for any covariates) using a t-test to compare the means of the two groups.

```{r}
t.test(ecls$c5r2mtsc_std ~ ecls$catholic)
```

### Balance Table

Now, let's create a balance table to identify the relevant co-variates in the unmatched dataset.

```{r warning=FALSE, message=FALSE}
ecls %>%
  group_by(catholic) %>%
  summarise_all(funs(mean(., na.rm = T))) %>% # summarize the mean of every variable
  t() %>% # transpose data
  as.data.frame() %>% # after transposing, convert from matrix to df
  add_rownames("variable") %>% # rownames to explicit column
  rename(no_catholic = V1, catholic = V2) %>% # rename columns that are created by as.data.frame
  mutate(difference = catholic - no_catholic,
         differencePerc = difference / (catholic + no_catholic)) %>% # create new columns for difference between groups
  mutate_if(is.numeric, round, 3) %>% # round numeric columns
  kable() %>% # create kable table
  kable_styling() # view kable table
```

We can see that the largest magnitude differences are for:

- variables on race
- w3momed_hsb: dummy variable with 1 = mother's education level is high school or below, 0 = some college or above
- w3income: family income
- w3povrty: poverty score dummy
- p5fstamp: food stamp dummy


### Exploring MatchIt

MatchIt is designed for causal inference with a dichotomous treatment variable and a set of pretreatment control variables. Any number or type of dependent variables can be used. 

The basic syntax is as follows: <br>
*matched <- matchit(treat ~ x1 + x2, data = mydata)*

MatchIt is capable of using several matching methods:

- Exact (*method = "exact"*): The simplest version of matching is exact. This technique matches each treated unit to all possible control units with exactly the same values on all the covariates, forming subclasses such that within each subclass all units (treatment and control) have the same covariate values. 
- Subclassification (*method = "subclass"*): When there are many covariates (or some covariates can take a large number of values),
finding sufficient exact matches will often be impossible. The goal of subclassification is to form subclasses, such that in each the distribution (rather than the exact values) of covariates for the treated and control groups are as similar as possible.
- Nearest Neighbor (*method = "nearest"*): Nearest neighbor matching selects the best control matches for each individual
in the treatment group. Matching is done using a distance measure (propensity score) specified by the distance option (default = logit).
- As well as optimal matching, full matching, genetic matching, and coarsened exact matching, all of which are detailed in the documentation.

A few additional arguments are important to know about:

- *distance*: this refers to propensity scores. There are many options for how to calculate these within MatchIt.
- *discard*: specifies whether to discard units that fall outside some measure of support of the distance measure (default is "none", discard no units). For example, if some treated units have extremely high propensity scores that are higher than any control units, we could drop those.
- *replace*: logical value indicating whether each control unit can be matched to more than one treated unit (default is *replace = FALSE*, each control unit is used at most once).
- *ratio*: the number of control units to match to each treated unit (default = 1).
- There are also some optional arguments for most of the matching methods, which you can read about in the documentation if you are interested.

### Exact Matching

We can use a combination of the results from our balance table and theory to identify which variables to use for matching. Let's perform an exact match with:

- race_white: Is the student white (1) or not (0)?
- p5hmage: Mother’s age
- w3income: Family income
- p5numpla: Number of places the student has lived for at least 4 months
- w3momed_hsb: Is the mother’s education level high-school or below (1) or some college or more (0)?

```{r}
# first we must omit missing values (MatchIt does not allow missings)
match_data <- ecls %>% 
  select(catholic, c5r2mtsc_std, race_white, p5hmage, 
         w3income, p5numpla, w3momed_hsb) %>% 
  na.omit()

# perform exact match
exact_match <- matchit(catholic ~ race_white + p5hmage + w3income +
                         p5numpla + w3momed_hsb, 
                       method = "exact", 
                       data = match_data)

# grab the matched data
data_exact_match <- match.data(exact_match)

# estimate t-test again
t.test(data_exact_match$c5r2mtsc_std ~ data_exact_match$catholic)
```

Now we can see that the mean in the group that did not attend Catholic school is actually about 0.10 higher than the mean for those who did. The results are statistically significant given that the confidence interval does not contain zero, and we have a lot p-value.

### Propensity Scores

If we want to perform non-exact matching, we need to estimate propensity scores. We can do this using logistic regression on the unmatched data set.

```{r, results = 'asis'}
# create a new column with income by the thousands for more interpretable output
ecls <- ecls %>% 
  mutate(w3income_1k = w3income / 1000) 

# estimate logit model
m_ps <- glm(catholic ~ race_white + w3income_1k + 
              p5hmage + p5numpla + w3momed_hsb,
            family = binomial(), 
            data = ecls)

# use the stargazer package to view the output
# NOTE: you MUST include results = "asis" in the chunk header for this to be visible once knitted
stargazer(m_ps, type = "html")

# extract predicted probabilities
# type = "response" option tells R to output probabilities of the form P(Y = 1|X)
prs_df <- data.frame(pr_score = predict(m_ps, type = "response"),
                     catholic = m_ps$model$catholic) # the actual values
```

Let's plot the propensity scores by treatment group to explore common support:

```{r warning=FALSE, message=FALSE}
prs_df %>%
  ggplot(aes(x = pr_score, fill = factor(catholic))) +
  geom_density(alpha = 0.5) +
  labs(x = "Propensity Score Distribution: Treatment and Control Groups",
       fill = "Catholic School Attendance")
```

### Non-Exact Matching

We can now estimate one-to-one nearest neighbor matching with replacement using the propensity scores. Try it yourself! All the info you need to complete this is available in the Exploring Matching section. Scroll down for the solution when you're ready ;-)
<div style="margin-bottom:500px;">
</div>
```{r}
one_match <- matchit(catholic ~ race_white + w3income + p5hmage +
                       p5numpla + w3momed_hsb,
                     method = "nearest", 
                     ratio = 1, 
                     replace = TRUE,
                     data = match_data)

summary(one_match)
```

We can interpret the resulting output as follows:

- Summary of balance for all data: Comparison of the means for all the data without matching
- Summary of balance for matched data: Comparison of means for matched data. Looking for them to become similar.
- Percent balance improvement: Higher is better, close to 100 is ideal.
- Sample sizes: How many units were matched in the control/treatment groups. 

Now, let's plot the propensity scores for the treated and untreated units. 

```{r}
# simple plot - check out the cobalt package for fancier options, or use ggplot2 to create your own!
plot(one_match, type = "hist")
```

Try creating a balance table like the one we did before, just this time using the new data from our *one_match* matching. Scroll down for the answer when you're ready.
<div style="margin-bottom:500px;">
</div>
```{r warning=FALSE, message=FALSE}
# grab data set
data_prop_match <- match.data(one_match)

# check balance
data_prop_match %>%
  group_by(catholic) %>%
  summarise_all(funs(mean)) %>%
  t() %>% 
  as.data.frame() %>% 
  add_rownames("variable") %>% 
  rename(no_catholic = V1, catholic = V2) %>% 
  mutate(difference = catholic - no_catholic,
         differencePerc = difference / (catholic + no_catholic)) %>% 
  mutate_if(is.numeric, round, 3) %>% 
  kable() %>% 
  kable_styling()
```

Those means look very close. Hooray.

Finally, estimate the treatment effect on the matched data set:
<div style="margin-bottom:500px;">
</div>
```{r}
t.test(data_prop_match$c5r2mtsc_std ~ data_prop_match$catholic)
```

As with the exact matching, we can see that those that did not attend Catholic school performed better on the test than those who did, and the results are statistically significant.

We can see the direction and magnitude of the covariate effects using a simple linear model:

```{r, results = 'asis'}
lm_matched <- lm(c5r2mtsc_std ~ catholic + race_white + p5hmage +
                  I(w3income / 1000) + p5numpla + w3momed_hsb, 
                 data = data_prop_match)

# use the stargazer package to view the output
# NOTE: you MUST include results = "asis" in the chunk header for this to be visible once knitted
stargazer(lm_matched, type = "html")
```
